### 2025-04-18 paper.tex
chore: Draft paper.

\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{listings}
\usepackage{xcolor}
\usepackage{geometry}
\usepackage{parskip}
\usepackage{amsmath}
\usepackage{url}

\geometry{
 letterpaper,
 left=20mm,
 top=20mm,
 }

\graphicspath{{images/}}

\title{\vspace{-2cm}STA5176 Class Project}
\author{Farooq Mahmud}
\date{April 21, 2024}

\definecolor{backcolor}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolor},
    captionpos=b                  
}

\lstset{style=mystyle}

\begin{document}

\maketitle
\tableofcontents

\section{Introduction}
The objective of the class project is to analyze a dataset containing variables related to Alzheimer's Disease mortality\cite{amin2018geographic} using SAS \cite{sasHelpCenter}. The analysis is conducted using various statistical methods learned in class.

\subsection{Variable of interest}
Each team member selected one variable to serve as the focus of their analysis. In this report, the variable of interest is diabetes. Most of the statistical methods shown in this report will be performed on the diabetes variable. Additional variables will be analyzed in conjunction with the diabetes variable when needed, \textit{e.g} for linear regression.

As a first step, the frequency counts by state were generated (Figure \ref{fig:state-freq}) using the SAS program in Listing \ref{lst:listing-state-freq}. There is data for six states with most of the data belonging to Georgia, Texas, and Virginia. The original dataset contained data for all fifty states. There are 715 total observations in the pared down dataset.

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{images/state-freq.png}
    \caption{Frequency by state}
    \label{fig:state-freq}
\end{figure}

\begin{lstlisting}[language=SAS,caption=Frequency by state,captionpos=b,label=lst:listing-state-freq]
    proc freq data=proj.alz;
    tables state;
    run;
\end{lstlisting}

\subsection{Outliers}
Detecting outliers is important because it could indicate a problem with the data which can impact subsequent analyses. The program in Listing \ref{lst:outliers} outputs box plots for the diabetes variable, grouped by state.

The box plots in \ref{appendix:boxplots} show that there are outliers in Texas, Virginia, and Washington. While further analysis of the outliers are out of scope for the project, a next step would be to look at these values to ensure that the values did not result from data entry errors.

\begin{lstlisting}[language=SAS,caption=Box plots of the diabetes variable for each state,captionpos=b,label=lst:outliers]
    proc univariate plot data=proj.alz;
    by state;
    var diabetes; run;
\end{lstlisting}

\section{Statistical Methods}
The bulk of this report explains the statistical methods used to analyze the diabetes variable using SAS.

\subsection{Significance Level}
Unless otherwise specified, the level of significance, $\alpha$ is $0.05$.

\subsection{Testing for Normality}
Many of the statistical techniques used in the project assume that the data follows a normal distribution. Therefore, it is a good idea to verify that assumption. The SAS program in Listing \ref{lst:listing-norm-test} runs the Shapiro-Wilk test and generates a Q-Q plot for the diabetes variable. These items are shown in Table \ref{tab:norm-test}.

\begin{lstlisting}[language=SAS,caption=SAS progam to test for the normality of the diabetes variable,captionpos=b,label=lst:listing-norm-test]
    proc univariate plot normal data=proj.alz;
    var diabetes; 
    run;
\end{lstlisting}

\begin{table}[ht]
    \begin{tabular}{ll}
         \includegraphics[scale=0.7]{images/shapiro-wilk.png} & \includegraphics[scale=0.5]{images/q-q-plot.png} \\
    \end{tabular}
    \caption{Normality test results for the diabetes variable}
    \label{tab:norm-test}
\end{table}

\subsubsection{Interpretation} \label{normality-interpretation}
The $p$-value for the Shapiro-Wilk test is less than $\alpha$. Therefore, the null hypothesis that the data follows a normal distribution is rejected. The lack of normality can be seen in the Q-Q plot in Table \ref{tab:norm-test} in that the plot is more of a "snake" shape than a straight line.

The lack of normality in the diabetes variable, or any other variable, doesn't invalidate the use of the subsequent statistical techniques explained in this report. As there are 715 observations, the Central Limit Theorem allows us to safely assume a normal distribution \cite{Ott_Longnecker_2016}.

\subsection{Correlation}
The program in Listing \ref{lst:corr} outputs the Pearson correlation coefficients for all variables in the dataset.

\begin{lstlisting}[language=SAS,caption=SAS correlation program,captionpos=b,label=lst:corr]
    proc corr data=proj.alz;
    var mental_distress	smoking_rate physical_inactivity diabetes
    heart_disease cancer fine_pm_2_5 mercury_tpy lead_tpy poverty_percent;
    run;
\end{lstlisting}

The table of coefficients are shown in \ref{appendix:corr}.

\subsubsection{Interpretation}
Looking at the diabetes variable row in the table of correlation coefficients, there seems to be a relatively strong positive correlation between diabetes and smoking as well as diabetes and physical inactivity.

\section{Testing for Equal Means (Assuming Normality)}

The mean diabetes for each state appear to be very different for the six states as shown in the box plots in \ref{appendix:boxplots}. However, this is not a true statistical test. In order to properly test for equal means, an ANOVA $F$-test is performed. Each state is a group and the null hypothesis is that mean diabetes level for all states are equal. The program in Listing \ref{lst:anova} outputs the $F$-test results shown in Table \ref{tab:anova}.

\begin{lstlisting}[language=SAS,caption=ANOVA F-test SAS progam,captionpos=b,label=lst:anova]
    proc glm data=proj.alz;
    class state;
    model diabetes=state;
    run;
\end{lstlisting}

\begin{table}[ht]
    \begin{tabular}{ll}
         \includegraphics[scale=0.7]{images/f-test.png} & \includegraphics[scale=0.6]{images/f-test-boxplots.png} \\
    \end{tabular}
    \caption{Normality test results for the diabetes variable}
    \label{tab:anova}
\end{table}

\subsubsection{Interpretation}
The ANOVA $F$-test $p$-value is less than $\alpha$. Therefore the null hypothesis is rejected. The mean diabetes level for the states are not equal.

The side-by-side box plots visually show the differences in means. California has the lowest level of diabetes and Georgia the highest. Subsequent sections will apply statistical tests to analytically determine which states are statistically significantly different from each other.

\section{Testing for Equal Variances}
Two methods in SAS exist to test for equal variances amongst groups - Bartlett's test and the Brown-Forsythe-Levene (BFL) test.

Bartlett's test is not appropriate to use with this data because it assumes equal sample sizes. As can be seen in Figure \ref{fig:state-freq}, the sample sizes are not equal for each state. The BFL test is not limited by this constraint. The program in \ref{lst:bfl} runs the BFL-test. The test's output is in Figure \ref{fig:bfl}.

\begin{lstlisting}[language=SAS,caption=BFL test SAS progam,captionpos=b,label=lst:bfl]
    proc glm data=proj.alz;
    class state;
    model diabetes=state;
    means state/hovtest=bf;
    run;
\end{lstlisting}

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{images/bfl.png}
    \caption{BFL test results}
    \label{fig:bfl}
\end{figure}

\subsection{Interpretation}
The $p$-value for the BFL test statistic is less than $\alpha$. Therefore, the null hypothesis that the states have equal variances is rejected. The reader can reconcile this with the side-by-side box plots in Table \ref{tab:anova}.

\section{Testing for Equal Means (Assuming non-Normality)}
As stated in \ref{normality-interpretation}, the Central Limit Theorem allows the diabetes data to be treated as if it were normally distributed. However methods do exist to test for equal means when the data is not normal. One such method is the Kruskal-Wallis test.

\subsection{Kruskal-Wallis Test}
The null hypothesis of the Kruskal-Wallis (KW) test is that all \textit{distributions}, \textit{i.e.} means and variances are equal. The code in Listing \ref{lst:kw} runs the KW test. The program's output is shown in Figure \ref{fig:kw}.

\begin{lstlisting}[language=SAS,caption=Kruskal-Wallis test SAS progam,captionpos=b,label=lst:kw]
    proc npar1way wilcoxon data=proj.alz;
    class state;
    var diabetes;
    run;
\end{lstlisting}

\begin{figure}[ht]
    \centering
    \includegraphics[scale=1]{images/kw.png}
    \caption{Kruskal-Wallis test result}
    \label{fig:kw}
\end{figure}

\subsubsection{Interpretation}
The $p$-value for the KW test statistic is less than $\alpha$. Therefore the null hypothesis that all distributions are equal is rejected. This result reconciles with the results of the ANOVA $F$-test (all means are not equal) and the BFL test (all variances are not equal).

\section{Multiple Comparisons}
The statistical tests done so far have shown that the mean level of diabetes amongst the states are not equal. This begs the question - "which states are significantly different from each other?" Multiple comparison tests helps answer this question.

SAS supports several such tests as shown in Table \ref{tab:mc-tests}.

\begin{table}[ht]
    \centering
    \begin{tabular}{|c|c|c|}
    \hline
        Fisher's Least Significant Difference (LSD) & Bonferroni & Sidak \\
        \hline
         Tukey & Student-Newman-Keuls (SNK) & Duncan \\
         \hline
    \end{tabular}
    \caption{Multiple comparisons tests}
    \label{tab:mc-tests}
\end{table}

Listing \ref{lst:mc} shows the SAS program that runs all of the aforementioned multiple comparison tests. The test results are shown in \ref{appendix:mc-diabetes}.

\begin{lstlisting}[language=SAS,caption=Multiple comparison tests SAS progam,captionpos=b,label=lst:mc]
    proc glm data=proj.alz;
    class state;
    model diabetes=state;
    means state/lsd bon sidak tukey snk duncan lines;
    run;
\end{lstlisting}

\subsection{Interpretation}
States connected by the same line are not statistically significantly different at the pre-determined $\alpha$ level of $0.05$. All the tests show that Florida and Virginia's mean diabetes levels are not significantly different. The box plots visually show that the states' mean diabetes levels are close. The remaining states are all significantly different from every other state. As can be gleaned from the box plots, Georgia and California have the highest and lowest mean diabetes levels, respectively.

Although all tests gave the same result, this is not always the case. Running the program in Listing \ref{lst:mc} with the cancer variable gives the result shown in \ref{appendix:mc-cancer}. Fisher's LSD, SNK, and Duncan's tests have the same result. Bonferroni's, Sidak, and Tukey's tests have the same result but the groups differ from the other three tests.

\subsection{Contrasts}
Another interesting question to ask is "Does the mean diabetes level significantly differ between west coast and east coast states?" Contrasts can help answer this question using statistical tests.

Assume California and Washington are west coast states. Assume Florida, Georgia, and Virginia are east coast states. The SAS program in Listing \ref{lst:contrasts-coast} runs this test.

\begin{lstlisting}[language=SAS,caption=Contrast SAS program comparing west coast and east coast states,captionpos=b,label=lst:contrasts-coast]
    proc sort data=proj.alz;
    by state;
    run;

    proc glm data=proj.alz;
    class state;
    model diabetes=state;
    contrast 'West coast vs. East coast states' state -3 2 2 0 2 -3;
    run;
\end{lstlisting}

The test result is shown in Figure \ref{fig:mc-coast-result}.

\begin{figure}[ht]
    \centering
    \includegraphics[scale=1]{images/multiple-comparisons/contrast-coasts.png}
    \caption{West coast vs. east coast constrast test result}
    \label{fig:mc-coast-result}
\end{figure}

\subsubsection{Interpretation}
The $p$-value of the test statistic is less than $\alpha$. Therefore, the null hypothesis that the mean diabetes level for west coast and east coast states are equal is rejected.

\section{Analyzing Categorical Data}
Suppose the diabetes variable is classified into four categories - "Low", "Medium", "High" and "Severe". A $\chi^2$ test for homogeneity can be carried out to determine if the four groups within each state are homogeneous.

Listing \ref{lst:severity} creates a new column named \textbf{severity} in the \textbf{proj.alz} dataset.

\begin{lstlisting}[language=SAS,caption=Adding the severity column to the dataset,captionpos=b,label=lst:severity]
    data proj.alz; 
    set proj.alz; 
    if diabetes le 8.76 then severity='Low'; 
    else if diabetes gt 8.76 and diabetes le 9.4 then severity='Medium'; 
    else if diabetes gt 9.4 and diabetes le 10.62 then severity='High'; 
    else if diabetes gt 10.62 then severity='Severe'; 
    run;
\end{lstlisting}

The numeric range of the severity levels are based on the quartiles which were obtained by running the SAS program in Listing \ref{lst:quartiles}.

\begin{lstlisting}[language=SAS,caption=Obtaining quartiles for the diabetes variable,captionpos=b,label=lst:quartiles]
    proc univariate data=proj.alz;
    var diabetes;
    run;
\end{lstlisting}

Finally, the $\chi^2$ test is run using the code in Listing \ref{lst:chisq}.

\begin{lstlisting}[language=SAS,caption=$\chi^2$ test for homogeneity variable,captionpos=b,label=lst:chisq]
    proc freq data=proj.alz;
    table severity*state/chisq;
    run;
\end{lstlisting}

The test results are shown in Figure \ref{fig:chisq-results}. The frequency table of severity levels by state is shown in Figure \ref{fig:chisq-freq}.

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{images/multiple-comparisons/chisq-results.png}
    \caption{$\chi^2$ test results}
    \label{fig:chisq-results}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{images/multiple-comparisons/chisq-freq.png}
    \caption{$\chi^2$ frequency table}
    \label{fig:chisq-freq}
\end{figure}

\clearpage

\subsection{Interpretation}
The $p$-value for the $\chi^2$ test statistic is less than $\alpha$. Therefore, the null hypothesis that the groups are homogeneous is rejected. The frequency table shows marked differences between the groups. For instance, California and Washington have very few instances of severe and high levels of diabetes. The opposite is true for Georgia and Florida.

\section{Linear Regression}
Linear regression helps answer the question - "Is there a strong relationship between diabetes and physical inactivity?" Listing \ref{lst:reg} shows the SAS program that runs the regression.

\begin{lstlisting}[language=SAS,caption=Linear regression SAS program variable,captionpos=b,label=lst:reg]
    proc reg data=proj.alz;
    model diabetes=physical_inactivity/clb;
    run;
\end{lstlisting}

\subsection{Parameter estimates}
The parameter estimates for the slope and intercept are shown in Figure \ref{fig:reg-param-est}.

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{images/regression/reg-param-est.png}
    \caption{Regression parameter estimates}
    \label{fig:reg-param-est}
\end{figure}

\subsubsection{Interpretation}
The regression equation is \begin{math} \text{diabetes}=3.0275+0.26031(\text{physical\_inactivity}) \end{math}. This means that as physical inactivity increases, so does the diabetes level. That is all that can be said. In order to make sense of the literal numeric values, more information about what the numbers represent, including units, need to be understood.

The output also shows the 95\% confidence limits for the slope and intercept values. Lastly, the $p$-values denote the probability that the true slope and intercept are zero. Since the $p$-values for both are less than $\alpha$, the null hypotheses that the true slope and intercept are zero, is rejected.

Figure \ref{fig:reg-fit-plot} shows the regression fit plot. This plot shows the positive correlation between physical inactivity and diabetes.

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.3]{images/regression/reg-fit-plot.png}
    \caption{Regression fit plot}
    \label{fig:reg-fit-plot}
\end{figure}

\subsection{Fit Metrics}
Figure \ref{fig:reg-r-sq} shows the $R^2$ value of the regression. The value means that about 52\% of the total variability in diabetes level is explained by the linear regression of diabetes level on physical inactivity levels. This is a mediocre fit. Note that the $R^2$ value is different from the Pearson correlation coefficient shown in \ref{appendix:corr} for diabetes vs. physical activity. The mean square error, $MS_W$ is an important fit metric that can be used in conjunction with $R^2$ to compare the strength of fit between multiple independent variables.

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{images/regression/reg-r-sq-mse.png}
    \caption{Fit metrics for physical inactivity vs. diabetes}
    \label{fig:reg-r-sq}
\end{figure}

The code in Listing \ref{lst:reg-multiple} runs the regression between (diabetes, smoking rate) and (diabetes, poverty percent). The fit metrics $R^2$ and $MS_W$ can be compared to determine which variable provides the best fit. \ref{tab:reg-multiple} shows that physical inactivity provides the best fit to the data as it has the highest $R_2$ and lowest $MS_W$.

\begin{lstlisting}[language=SAS,caption=Linear regression SAS program for additional variables,captionpos=b,label=lst:reg-multiple]
    proc reg data=proj.alz;
    model diabetes=smoking_rate/clb;
    run;

    proc reg data=proj.alz;
    model diabetes=poverty_percent/clb;
    run;
\end{lstlisting}

\begin{table}[ht]
    \centering
    \begin{tabular}{|l|l|l|}
    \hline
         Variable & $R^2$  & $MS_W$ \\
         \hline
         physical inactivity& 0.5159  & 1.05256 \\
         \hline
         smoking rate & 	0.3817  & 1.34436 \\
         \hline
         poverty percent & 0.1718  & 1.80068 \\
         \hline
    \end{tabular}
    \caption{Fit statistics for the three variables}
    \label{tab:reg-multiple}
\end{table}

\clearpage

\section{Predictions}
$R_2$ and lowest $MS_W$ are not the best metrics to evaluate which variables are the best predictors of diabetes. Predicting means applying the model to new data as opposed to the existing data. Therefore, the PRESS metric is used instead. The code in Listing \ref{lst:reg-predict} outputs the PRESS metrics for physical inactivity, smoking rate, and poverty percent variables. The metrics are presented in Table \ref{tab:prediction}.

\begin{lstlisting}[language=SAS,caption=Linear regression SAS program for additional variables,captionpos=b,label=lst:reg-predict]
    proc reg data=proj.alz;
    model diabetes=physical_inactivity/cli clb p;
    run;

    proc reg data=proj.alz;
    model diabetes=smoking_rate/cli clb p;
    run;

    proc reg data=proj.proc reg data=proj.alz;;
    model diabetes=poverty_percent/cli clb p;
    run;
\end{lstlisting}

\begin{table}[ht]
    \centering
    \begin{tabular}{|l|l|l|}
    \hline
         Variable & Sum of squared residuals  & PRESS \\
         \hline
         physical inactivity& 746.26521  & 749.91380 \\
         \hline
         smoking rate & 	953.15029  & 958.52529 \\
         \hline
         poverty percent & 1276.67958  & 1284.90979 \\
         \hline
    \end{tabular}
    \caption{Prediction statistics for the three variables}
    \label{tab:prediction}
\end{table}

\subsection{Interpretation}
Physical inactivity is the best predictor of diabetes since it has the lowest PRESS value.

\clearpage

\section{Lack of Fit}
The data may be such that a linear regression model (as opposed to a polynomial model) is not appropriate. A lack of fit test is a statistical test that determines whether a linear regression model is appropriate. Listing \ref{lst:reg-lof} shows the SAS program used to run the lack of fit test.

\begin{lstlisting}[language=SAS,caption=SAS program for lack of fit test,captionpos=b,label=lst:reg-lof]
    proc reg data=proj.alz;
    model diabetes=physical_inactivity/lackfit;
    run;
\end{lstlisting}

Its output is shown in Figure \ref{fig:reg-lof}.

\subsection{Interpretation}
The $p$-value for the lack of fit test is greater than $\alpha$. Therefore, the null hypothesis that there is no lack of fit, is not rejected. A linear regression model seems appropriate.

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{images/regression/reg-lof.png}
    \caption{Regression lack of fit results}
    \label{fig:reg-lof}
\end{figure}

\clearpage

\section{Appendix} \label{appendix}

\subsection{Box plots of diabetes by state} \label{appendix:boxplots}
\begin{table}[ht]
\centering
    \begin{tabular}{|cc|cc|cc|}
    \hline
         California &  \includegraphics[scale=0.6]{images/boxplots/ca.png}  & Florida  & \includegraphics[scale=0.6]{images/boxplots/fl.png} & Georgia & \includegraphics[scale=0.6]{images/boxplots/ga.png} \\
         \hline
         Texas & \includegraphics[scale=0.6]{images/boxplots/tx.png} & Virginia  & \includegraphics[scale=0.6]{images/boxplots/va.png} & Washington  & \includegraphics[scale=0.6]{images/boxplots/wa.png} \\
         \hline
    \end{tabular}

\end{table}

\clearpage

\subsection{Correlation coefficients} \label{appendix:corr}

\begin{figure}[ht]
    \centering
    \includegraphics[scale=0.7]{images/proc-corr.png}

\end{figure}

\clearpage

\subsection{Multiple comparisons test results for the diabetes variable} \label{appendix:mc-diabetes}

\begin{table}[ht]
    \begin{tabular}{|ll|ll|}
    \hline
         Fisher's LSD & \includegraphics[scale=0.6]{images/multiple-comparisons/lsd.png} & Bonferroni & \includegraphics[scale=0.6]{images/multiple-comparisons/bon.png} \\
         \hline
         Sidak & \includegraphics[scale=0.6]{images/multiple-comparisons/sidak.png}  & Tukey  & \includegraphics[scale=0.6]{images/multiple-comparisons/tukey.png} \\
         \hline
         SNK & \includegraphics[scale=0.6]{images/multiple-comparisons/snk.png}  & Duncan & \includegraphics[scale=0.6]{images/multiple-comparisons/duncan.png} \\
         \hline
    \end{tabular}

\end{table}

\clearpage

\subsection{Multiple comparisons test results for the cancer variable} \label{appendix:mc-cancer}

\begin{table}[ht]
    \begin{tabular}{|ll|ll|}
    \hline
         Fisher's LSD & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/lsd.png} & Bonferroni & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/bon.png} \\
         \hline
         Sidak & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/sidak.png}  & Tukey  & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/tukey.png} \\
         \hline
         SNK & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/snk.png}  & Duncan & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/duncan.png} \\
         \hline
    \end{tabular}

\end{table}

\clearpage

\bibliographystyle{plain}
\bibliography{refs}

\end{document}

### 2025-04-18 paper.tex
chore: Draft paper.

@@
 
-\title{\vspace{-2cm}STA5176 Class Project}
+\title{\vspace{-2cm}Temporal Congestion Analysis and Predictive Modeling of Ride-Hailing Trips in NYC Using High-Volume FHV Data}
 \author{Farooq Mahmud}
-\date{April 21, 2024}
+\date{April 25, 2025}
 
@@
 \maketitle
-\tableofcontents
+
+\section{Abstract}
+This paper explores traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services like Uber and Lyft.
+Using data from all 12 months of 2024, we investigate how trip durations and speeds vary by time of day, day of week, and provider.
+After extensive preprocessing in PySpark, we identify temporal trends and build predictive models in R using linear and logistic regression.
+The study demonstrates that the dataset size and analysis complexity necessitate big data tools for processing and modeling at scale.
 
 \section{Introduction}
-The objective of the class project is to analyze a dataset containing variables related to Alzheimer's Disease mortality\cite{amin2018geographic} using SAS \cite{sasHelpCenter}.
-The analysis is conducted using various statistical methods learned in class.
-
-\subsection{Variable of interest}
-Each team member selected one variable to serve as the focus of their analysis.
-In this report, the variable of interest is diabetes.
-Most of the statistical methods shown in this report will be performed on the diabetes variable.
-Additional variables will be analyzed in conjunction with the diabetes variable when needed, \textit{e.g} for linear regression.
-
-As a first step, the frequency counts by state were generated (Figure \ref{fig:state-freq}) using the SAS program in Listing \ref{lst:listing-state-freq}.
-There is data for six states with most of the data belonging to Georgia, Texas, and Virginia.
-The original dataset contained data for all fifty states.
-There are 715 total observations in the pared down dataset.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.7]{images/state-freq.png}
-    \caption{Frequency by state}
-    \label{fig:state-freq}
-\end{figure}
-
-\begin{lstlisting}[language=SAS,caption=Frequency by state,captionpos=b,label=lst:listing-state-freq]
-    proc freq data=proj.alz;
-    tables state;
-    run;
-\end{lstlisting}
-
-\subsection{Outliers}
-Detecting outliers is important because it could indicate a problem with the data which can impact subsequent analyses.
-The program in Listing \ref{lst:outliers} outputs box plots for the diabetes variable, grouped by state.
-
-The box plots in \ref{appendix:boxplots} show that there are outliers in Texas, Virginia, and Washington.
-While further analysis of the outliers are out of scope for the project, a next step would be to look at these values to ensure that the values did not result from data entry errors.
-
-\begin{lstlisting}[language=SAS,caption=Box plots of the diabetes variable for each state,captionpos=b,label=lst:outliers]
-    proc univariate plot data=proj.alz;
-    by state;
-    var diabetes; run;
-\end{lstlisting}
-
-\section{Statistical Methods}
-The bulk of this report explains the statistical methods used to analyze the diabetes variable using SAS.
-
-\subsection{Significance Level}
-Unless otherwise specified, the level of significance, $\alpha$ is $0.05$.
-
-\subsection{Testing for Normality}
-Many of the statistical techniques used in the project assume that the data follows a normal distribution.
-Therefore, it is a good idea to verify that assumption.
-The SAS program in Listing \ref{lst:listing-norm-test} runs the Shapiro-Wilk test and generates a Q-Q plot for the diabetes variable.
-These items are shown in Table \ref{tab:norm-test}.
-
-\begin{lstlisting}[language=SAS,caption=SAS progam to test for the normality of the diabetes variable,captionpos=b,label=lst:listing-norm-test]
-    proc univariate plot normal data=proj.alz;
-    var diabetes; 
-    run;
-\end{lstlisting}
-
-\begin{table}[ht]
-    \begin{tabular}{ll}
-         \includegraphics[scale=0.7]{images/shapiro-wilk.png} & \includegraphics[scale=0.5]{images/q-q-plot.png} \\
-    \end{tabular}
-    \caption{Normality test results for the diabetes variable}
-    \label{tab:norm-test}
-\end{table}
-
-\subsubsection{Interpretation} \label{normality-interpretation}
-The $p$-value for the Shapiro-Wilk test is less than $\alpha$.
-Therefore, the null hypothesis that the data follows a normal distribution is rejected.
-The lack of normality can be seen in the Q-Q plot in Table \ref{tab:norm-test} in that the plot is more of a "snake" shape than a straight line.
-
-The lack of normality in the diabetes variable, or any other variable, doesn't invalidate the use of the subsequent statistical techniques explained in this report.
-As there are 715 observations, the Central Limit Theorem allows us to safely assume a normal distribution \cite{Ott_Longnecker_2016}.
-
-\subsection{Correlation}
-The program in Listing \ref{lst:corr} outputs the Pearson correlation coefficients for all variables in the dataset.
-
-\begin{lstlisting}[language=SAS,caption=SAS correlation program,captionpos=b,label=lst:corr]
-    proc corr data=proj.alz;
-    var mental_distress	smoking_rate physical_inactivity diabetes
-    heart_disease cancer fine_pm_2_5 mercury_tpy lead_tpy poverty_percent;
-    run;
-\end{lstlisting}
-
-The table of coefficients are shown in \ref{appendix:corr}.
-
-\subsubsection{Interpretation}
-Looking at the diabetes variable row in the table of correlation coefficients, there seems to be a relatively strong positive correlation between diabetes and smoking as well as diabetes and physical inactivity.
+Urban traffic congestion continues to be a pervasive problem affecting the efficiency of transportation systems, contributing to environmental concerns, and impacting commuter well-being.
+In densely populated cities like New York, where ride-hailing services such as Uber and Lyft operate at massive scale, understanding when and where congestion occurs is critical for optimizing urban mobility.
+This project investigates temporal patterns in congestion using High-Volume For-Hire Vehicle (FHV) trip data collected and published by the NYC Taxi and Limousine Commission (TLC).
+With over 200 million trip records from 2024 alone, the dataset provides a rich source of behavioral and operational insights.
+The central research question is: How do trip duration and speed, as proxies for congestion, vary across different times of day, days of the week, and ride-hailing providers?
+The objectives of the study are to (1) identify temporal congestion trends through descriptive analysis, and (2) develop regression models to predict trip duration or classify trips as congested based on their features.
+Given the volume and granularity of the dataset, traditional analysis tools fall short, necessitating the use of big data processing technologies like Apache Spark.
+This research aims to produce interpretable insights that can inform urban planning, traffic policy, and ride-hailing operations.
 

### 2025-04-18 paper.tex
chore: Draft paper.

@@
 
-\section{Testing for Equal Means (Assuming Normality)}
+\section{Background Review}
 
-The mean diabetes for each state appear to be very different for the six states as shown in the box plots in \ref{appendix:boxplots}.
-However, this is not a true statistical test.
-In order to properly test for equal means, an ANOVA $F$-test is performed.
-Each state is a group and the null hypothesis is that mean diabetes level for all states are equal.
-The program in Listing \ref{lst:anova} outputs the $F$-test results shown in Table \ref{tab:anova}.
+A growing body of research has explored the use of ride-hailing data for understanding urban transportation dynamics.
+Previous studies have focused on topics such as demand prediction, ride-sharing optimization, pricing strategy, and transportation equity.
+NYC’s Taxi and Limousine Commission has provided open access to large-scale trip data, which has enabled a range of analyses from temporal ridership patterns to spatial clustering of pick-up and drop-off zones \cite{nyctlc2023}.
+Several works have leveraged this dataset to evaluate ride-hailing’s impact on traffic flow and public transit usage.
+For example, a study by NYU Tandon found that New York City’s congestion surcharge reduced the number of ride-hailing trips but did not significantly alleviate congestion \cite{nyu2023}.
+Additionally, research from the CUNY Graduate Center examined the 2019 congestion surcharge’s effects on FHV usage in Manhattan, showing a decline in rides from charged zones \cite{cuny2023}.
+Other studies have investigated surge pricing’s influence on driver behavior \cite{surgepricing2022} and analyzed pay disparities across platforms using data from 2019 to 2022 \cite{ucla2023}.
+However, many of these analyses are limited in scope—often examining only a few months of data, focusing exclusively on yellow taxi or green cab datasets, or lacking predictive components.
+There remains a gap in literature concerning sustained, full-year temporal analysis of congestion based specifically on High-Volume FHV data.
+Furthermore, predictive modeling using such data is less common, particularly models focused on estimating congestion or trip delays.
+Our project addresses these shortcomings by utilizing a complete 12-month dataset and incorporating regression modeling.
+Additionally, our use of big data processing tools distinguishes this study from prior work that relies on single-machine or static data approaches.
 
-\begin{lstlisting}[language=SAS,caption=ANOVA F-test SAS progam,captionpos=b,label=lst:anova]
-    proc glm data=proj.alz;
-    class state;
-    model diabetes=state;
-    run;
-\end{lstlisting}
-
-\begin{table}[ht]
-    \begin{tabular}{ll}
-         \includegraphics[scale=0.7]{images/f-test.png} & \includegraphics[scale=0.6]{images/f-test-boxplots.png} \\
-    \end{tabular}
-    \caption{Normality test results for the diabetes variable}
-    \label{tab:anova}
-\end{table}
-
-\subsubsection{Interpretation}
-The ANOVA $F$-test $p$-value is less than $\alpha$. Therefore the null hypothesis is rejected. The mean diabetes level for the states are not equal.
-
-The side-by-side box plots visually show the differences in means.
-California has the lowest level of diabetes and Georgia the highest.
-Subsequent sections will apply statistical tests to analytically determine which states are statistically significantly different from each other.
-
-\section{Testing for Equal Variances}
-Two methods in SAS exist to test for equal variances amongst groups - Bartlett's test and the Brown-Forsythe-Levene (BFL) test.
-
-Bartlett's test is not appropriate to use with this data because it assumes equal sample sizes.
-As can be seen in Figure \ref{fig:state-freq}, the sample sizes are not equal for each state.
-The BFL test is not limited by this constraint.
-The program in \ref{lst:bfl} runs the BFL-test.
-The test's output is in Figure \ref{fig:bfl}.
-
-\begin{lstlisting}[language=SAS,caption=BFL test SAS progam,captionpos=b,label=lst:bfl]
-    proc glm data=proj.alz;
-    class state;
-    model diabetes=state;
-    means state/hovtest=bf;
-    run;
-\end{lstlisting}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.7]{images/bfl.png}
-    \caption{BFL test results}
-    \label{fig:bfl}
-\end{figure}
-
-\subsection{Interpretation}
-The $p$-value for the BFL test statistic is less than $\alpha$.
-Therefore, the null hypothesis that the states have equal variances is rejected.
-The reader can reconcile this with the side-by-side box plots in Table \ref{tab:anova}.
-
-\section{Testing for Equal Means (Assuming non-Normality)}
-As stated in \ref{normality-interpretation}, the Central Limit Theorem allows the diabetes data to be treated as if it were normally distributed.
-However methods do exist to test for equal means when the data is not normal.
-One such method is the Kruskal-Wallis test.
-
-\subsection{Kruskal-Wallis Test}
-The null hypothesis of the Kruskal-Wallis (KW) test is that all \textit{distributions}, \textit{i.e.} means and variances are equal.
-The code in Listing \ref{lst:kw} runs the KW test.
-The program's output is shown in Figure \ref{fig:kw}.
-
-\begin{lstlisting}[language=SAS,caption=Kruskal-Wallis test SAS progam,captionpos=b,label=lst:kw]
-    proc npar1way wilcoxon data=proj.alz;
-    class state;
-    var diabetes;
-    run;
-\end{lstlisting}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=1]{images/kw.png}
-    \caption{Kruskal-Wallis test result}
-    \label{fig:kw}
-\end{figure}
-
-\subsubsection{Interpretation}
-The $p$-value for the KW test statistic is less than $\alpha$.
-Therefore the null hypothesis that all distributions are equal is rejected.
-This result reconciles with the results of the ANOVA $F$-test (all means are not equal) and the BFL test (all variances are not equal).
-
-\section{Multiple Comparisons}
-The statistical tests done so far have shown that the mean level of diabetes amongst the states are not equal.
-This begs the question - "which states are significantly different from each other?" Multiple comparison tests helps answer this question.
-
-SAS supports several such tests as shown in Table \ref{tab:mc-tests}.
-
-\begin{table}[ht]
-    \centering
-    \begin{tabular}{|c|c|c|}
-    \hline
-        Fisher's Least Significant Difference (LSD) & Bonferroni & Sidak \\
-        \hline
-         Tukey & Student-Newman-Keuls (SNK) & Duncan \\
-         \hline
-    \end{tabular}
-    \caption{Multiple comparisons tests}
-    \label{tab:mc-tests}
-\end{table}
-
-Listing \ref{lst:mc} shows the SAS program that runs all of the aforementioned multiple comparison tests. The test results are shown in \ref{appendix:mc-diabetes}.
-
-\begin{lstlisting}[language=SAS,caption=Multiple comparison tests SAS progam,captionpos=b,label=lst:mc]
-    proc glm data=proj.alz;
-    class state;
-    model diabetes=state;
-    means state/lsd bon sidak tukey snk duncan lines;
-    run;
-\end{lstlisting}
-
-\subsection{Interpretation}
-States connected by the same line are not statistically significantly different at the pre-determined $\alpha$ level of $0.05$.
-All the tests show that Florida and Virginia's mean diabetes levels are not significantly different.
-The box plots visually show that the states' mean diabetes levels are close.
-The remaining states are all significantly different from every other state.
-As can be gleaned from the box plots, Georgia and California have the highest and lowest mean diabetes levels, respectively.
-
-Although all tests gave the same result, this is not always the case.
-Running the program in Listing \ref{lst:mc} with the cancer variable gives the result shown in \ref{appendix:mc-cancer}.
-Fisher's LSD, SNK, and Duncan's tests have the same result.
-Bonferroni's, Sidak, and Tukey's tests have the same result but the groups differ from the other three tests.
-
-\subsection{Contrasts}
-Another interesting question to ask is "Does the mean diabetes level significantly differ between west coast and east coast states?" Contrasts can help answer this question using statistical tests.
-
-Assume California and Washington are west coast states. Assume Florida, Georgia, and Virginia are east coast states. The SAS program in Listing \ref{lst:contrasts-coast} runs this test.
-
-\begin{lstlisting}[language=SAS,caption=Contrast SAS program comparing west coast and east coast states,captionpos=b,label=lst:contrasts-coast]
-    proc sort data=proj.alz;
-    by state;
-    run;
-
-    proc glm data=proj.alz;
-    class state;
-    model diabetes=state;
-    contrast 'West coast vs. East coast states' state -3 2 2 0 2 -3;
-    run;
-\end{lstlisting}
-
-The test result is shown in Figure \ref{fig:mc-coast-result}.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=1]{images/multiple-comparisons/contrast-coasts.png}
-    \caption{West coast vs. east coast constrast test result}
-    \label{fig:mc-coast-result}
-\end{figure}
-
-\subsubsection{Interpretation}
-The $p$-value of the test statistic is less than $\alpha$. Therefore, the null hypothesis that the mean diabetes level for west coast and east coast states are equal is rejected.
-
-\section{Analyzing Categorical Data}
-Suppose the diabetes variable is classified into four categories - "Low", "Medium", "High" and "Severe".
-A $\chi^2$ test for homogeneity can be carried out to determine if the four groups within each state are homogeneous.
-
-Listing \ref{lst:severity} creates a new column named \textbf{severity} in the \textbf{proj.alz} dataset.
-
-\begin{lstlisting}[language=SAS,caption=Adding the severity column to the dataset,captionpos=b,label=lst:severity]
-    data proj.alz; 
-    set proj.alz; 
-    if diabetes le 8.76 then severity='Low'; 
-    else if diabetes gt 8.76 and diabetes le 9.4 then severity='Medium'; 
-    else if diabetes gt 9.4 and diabetes le 10.62 then severity='High'; 
-    else if diabetes gt 10.62 then severity='Severe'; 
-    run;
-\end{lstlisting}
-
-The numeric range of the severity levels are based on the quartiles which were obtained by running the SAS program in Listing \ref{lst:quartiles}.
-
-\begin{lstlisting}[language=SAS,caption=Obtaining quartiles for the diabetes variable,captionpos=b,label=lst:quartiles]
-    proc univariate data=proj.alz;
-    var diabetes;
-    run;
-\end{lstlisting}
-
-Finally, the $\chi^2$ test is run using the code in Listing \ref{lst:chisq}.
-
-\begin{lstlisting}[language=SAS,caption=$\chi^2$ test for homogeneity variable,captionpos=b,label=lst:chisq]
-    proc freq data=proj.alz;
-    table severity*state/chisq;
-    run;
-\end{lstlisting}
-
-The test results are shown in Figure \ref{fig:chisq-results}. The frequency table of severity levels by state is shown in Figure \ref{fig:chisq-freq}.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.7]{images/multiple-comparisons/chisq-results.png}
-    \caption{$\chi^2$ test results}
-    \label{fig:chisq-results}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.7]{images/multiple-comparisons/chisq-freq.png}
-    \caption{$\chi^2$ frequency table}
-    \label{fig:chisq-freq}
-\end{figure}
-
-\clearpage
-
-\subsection{Interpretation}
-The $p$-value for the $\chi^2$ test statistic is less than $\alpha$.
-Therefore, the null hypothesis that the groups are homogeneous is rejected.
-The frequency table shows marked differences between the groups.
-For instance, California and Washington have very few instances of severe and high levels of diabetes.
-The opposite is true for Georgia and Florida.
-
-\section{Linear Regression}
-Linear regression helps answer the question - "Is there a strong relationship between diabetes and physical inactivity?" Listing \ref{lst:reg} shows the SAS program that runs the regression.
-
-\begin{lstlisting}[language=SAS,caption=Linear regression SAS program variable,captionpos=b,label=lst:reg]
-    proc reg data=proj.alz;
-    model diabetes=physical_inactivity/clb;
-    run;
-\end{lstlisting}
-
-\subsection{Parameter estimates}
-The parameter estimates for the slope and intercept are shown in Figure \ref{fig:reg-param-est}.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.7]{images/regression/reg-param-est.png}
-    \caption{Regression parameter estimates}
-    \label{fig:reg-param-est}
-\end{figure}
-
-\subsubsection{Interpretation}
-The regression equation is \begin{math} \text{diabetes}=3.0275+0.26031(\text{physical\_inactivity}) \end{math}.
-This means that as physical inactivity increases, so does the diabetes level.
-That is all that can be said.
-In order to make sense of the literal numeric values, more information about what the numbers represent, including units, need to be understood.
-
-The output also shows the 95\% confidence limits for the slope and intercept values.
-Lastly, the $p$-values denote the probability that the true slope and intercept are zero.
-Since the $p$-values for both are less than $\alpha$, the null hypotheses that the true slope and intercept are zero, is rejected.
-
-Figure \ref{fig:reg-fit-plot} shows the regression fit plot. This plot shows the positive correlation between physical inactivity and diabetes.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.3]{images/regression/reg-fit-plot.png}
-    \caption{Regression fit plot}
-    \label{fig:reg-fit-plot}
-\end{figure}
-
-\subsection{Fit Metrics}
-Figure \ref{fig:reg-r-sq} shows the $R^2$ value of the regression.
-The value means that about 52\% of the total variability in diabetes level is explained by the linear regression of diabetes level on physical inactivity levels.
-This is a mediocre fit.
-Note that the $R^2$ value is different from the Pearson correlation coefficient shown in \ref{appendix:corr} for diabetes vs.
-physical activity.
-The mean square error, $MS_W$ is an important fit metric that can be used in conjunction with $R^2$ to compare the strength of fit between multiple independent variables.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.7]{images/regression/reg-r-sq-mse.png}
-    \caption{Fit metrics for physical inactivity vs. diabetes}
-    \label{fig:reg-r-sq}
-\end{figure}
-
-The code in Listing \ref{lst:reg-multiple} runs the regression between (diabetes, smoking rate) and (diabetes, poverty percent).
-The fit metrics $R^2$ and $MS_W$ can be compared to determine which variable provides the best fit.
-\ref{tab:reg-multiple} shows that physical inactivity provides the best fit to the data as it has the highest $R_2$ and lowest $MS_W$.
-
-\begin{lstlisting}[language=SAS,caption=Linear regression SAS program for additional variables,captionpos=b,label=lst:reg-multiple]
-    proc reg data=proj.alz;
-    model diabetes=smoking_rate/clb;
-    run;
-
-    proc reg data=proj.alz;
-    model diabetes=poverty_percent/clb;
-    run;
-\end{lstlisting}
-
-\begin{table}[ht]
-    \centering
-    \begin{tabular}{|l|l|l|}
-    \hline
-         Variable & $R^2$  & $MS_W$ \\
-         \hline
-         physical inactivity& 0.5159  & 1.05256 \\
-         \hline
-         smoking rate & 	0.3817  & 1.34436 \\
-         \hline
-         poverty percent & 0.1718  & 1.80068 \\
-         \hline
-    \end{tabular}
-    \caption{Fit statistics for the three variables}
-    \label{tab:reg-multiple}
-\end{table}
-
-\clearpage
-
-\section{Predictions}
-$R_2$ and lowest $MS_W$ are not the best metrics to evaluate which variables are the best predictors of diabetes.
-Predicting means applying the model to new data as opposed to the existing data.
-Therefore, the PRESS metric is used instead.
-The code in Listing \ref{lst:reg-predict} outputs the PRESS metrics for physical inactivity, smoking rate, and poverty percent variables.
-The metrics are presented in Table \ref{tab:prediction}.
-
-\begin{lstlisting}[language=SAS,caption=Linear regression SAS program for additional variables,captionpos=b,label=lst:reg-predict]
-    proc reg data=proj.alz;
-    model diabetes=physical_inactivity/cli clb p;
-    run;
-
-    proc reg data=proj.alz;
-    model diabetes=smoking_rate/cli clb p;
-    run;
-
-    proc reg data=proj.proc reg data=proj.alz;;
-    model diabetes=poverty_percent/cli clb p;
-    run;
-\end{lstlisting}
-
-\begin{table}[ht]
-    \centering
-    \begin{tabular}{|l|l|l|}
-    \hline
-         Variable & Sum of squared residuals  & PRESS \\
-         \hline
-         physical inactivity& 746.26521  & 749.91380 \\
-         \hline
-         smoking rate & 	953.15029  & 958.52529 \\
-         \hline
-         poverty percent & 1276.67958  & 1284.90979 \\
-         \hline
-    \end{tabular}
-    \caption{Prediction statistics for the three variables}
-    \label{tab:prediction}
-\end{table}
-
-\subsection{Interpretation}
-Physical inactivity is the best predictor of diabetes since it has the lowest PRESS value.
-
-\clearpage
-
-\section{Lack of Fit}
-The data may be such that a linear regression model (as opposed to a polynomial model) is not appropriate.
-A lack of fit test is a statistical test that determines whether a linear regression model is appropriate.
-Listing \ref{lst:reg-lof} shows the SAS program used to run the lack of fit test.
-
-\begin{lstlisting}[language=SAS,caption=SAS program for lack of fit test,captionpos=b,label=lst:reg-lof]
-    proc reg data=proj.alz;
-    model diabetes=physical_inactivity/lackfit;
-    run;
-\end{lstlisting}
-
-Its output is shown in Figure \ref{fig:reg-lof}.
-
-\subsection{Interpretation}
-The $p$-value for the lack of fit test is greater than $\alpha$. Therefore, the null hypothesis that there is no lack of fit, is not rejected. A linear regression model seems appropriate.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.7]{images/regression/reg-lof.png}
-    \caption{Regression lack of fit results}
-    \label{fig:reg-lof}
-\end{figure}
-
-\clearpage
-
-\section{Appendix} \label{appendix}
-
-\subsection{Box plots of diabetes by state} \label{appendix:boxplots}
-\begin{table}[ht]
-\centering
-    \begin{tabular}{|cc|cc|cc|}
-    \hline
-         California &  \includegraphics[scale=0.6]{images/boxplots/ca.png}  & Florida  & \includegraphics[scale=0.6]{images/boxplots/fl.png} & Georgia & \includegraphics[scale=0.6]{images/boxplots/ga.png} \\
-         \hline
-         Texas & \includegraphics[scale=0.6]{images/boxplots/tx.png} & Virginia  & \includegraphics[scale=0.6]{images/boxplots/va.png} & Washington  & \includegraphics[scale=0.6]{images/boxplots/wa.png} \\
-         \hline
-    \end{tabular}
-
-\end{table}
-
-\clearpage
-
-\subsection{Correlation coefficients} \label{appendix:corr}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.7]{images/proc-corr.png}
-
-\end{figure}
-
-\clearpage
-
-\subsection{Multiple comparisons test results for the diabetes variable} \label{appendix:mc-diabetes}
-
-\begin{table}[ht]
-    \begin{tabular}{|ll|ll|}
-    \hline
-         Fisher's LSD & \includegraphics[scale=0.6]{images/multiple-comparisons/lsd.png} & Bonferroni & \includegraphics[scale=0.6]{images/multiple-comparisons/bon.png} \\
-         \hline
-         Sidak & \includegraphics[scale=0.6]{images/multiple-comparisons/sidak.png}  & Tukey  & \includegraphics[scale=0.6]{images/multiple-comparisons/tukey.png} \\
-         \hline
-         SNK & \includegraphics[scale=0.6]{images/multiple-comparisons/snk.png}  & Duncan & \includegraphics[scale=0.6]{images/multiple-comparisons/duncan.png} \\
-         \hline
-    \end{tabular}
-
-\end{table}
-
-\clearpage
-
-\subsection{Multiple comparisons test results for the cancer variable} \label{appendix:mc-cancer}
-
-\begin{table}[ht]
-    \begin{tabular}{|ll|ll|}
-    \hline
-         Fisher's LSD & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/lsd.png} & Bonferroni & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/bon.png} \\
-         \hline
-         Sidak & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/sidak.png}  & Tukey  & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/tukey.png} \\
-         \hline
-         SNK & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/snk.png}  & Duncan & \includegraphics[scale=0.6]{images/multiple-comparisons-cancer/duncan.png} \\
-         \hline
-    \end{tabular}
-
-\end{table}
-
-\clearpage
+\section{Conclusion}
+This study highlights the feasibility and value of using big data tools to analyze ride-hailing trip data for congestion insights.
+It confirms the need for scalable data processing when working with multi-year FHV datasets and shows that simple predictive models can effectively estimate congestion levels using temporal features.
+Future work could include incorporating spatial features, extending modeling to multi-year data, or testing real-time streaming analysis.
 

### 2025-04-19 paper.tex

@@
 \section{Abstract}
-This paper explores traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services like Uber and Lyft.
+This paper explores traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services such as Uber and Lyft.
 Using data from all 12 months of 2024, we investigate how trip durations and speeds vary by time of day, day of week, and provider.
-After extensive preprocessing in PySpark, we identify temporal trends and build predictive models in R using linear and logistic regression.
+After extensive pre-processing in Apache Spark (PySpark), we identify temporal trends and build predictive models in R using linear and logistic regression.
 The study demonstrates that the dataset size and analysis complexity necessitate big data tools for processing and modeling at scale.
@@
 Several works have leveraged this dataset to evaluate ride-hailing’s impact on traffic flow and public transit usage.
-For example, a study by NYU Tandon found that New York City’s congestion surcharge reduced the number of ride-hailing trips but did not significantly alleviate congestion \cite{nyu2023}.
+For example, a study by NYU's Tandon School of Engineering found that New York City’s congestion surcharge reduced the number of ride-hailing trips but did not significantly alleviate congestion \cite{nyu2023}.
 Additionally, research from the CUNY Graduate Center examined the 2019 congestion surcharge’s effects on FHV usage in Manhattan, showing a decline in rides from charged zones \cite{cuny2023}.
@@
 Additionally, our use of big data processing tools distinguishes this study from prior work that relies on single-machine or static data approaches.
+
+\section{Method}
+This study follows a multi-phase approach consistent with the big data lifecycle, including data acquisition, pre-processing, analysis, and modeling.
+
+\subsection{Data Acquisition and Setup}
+We obtained the NYC TLC High-Volume FHV trip record dataset for all twelve months of 2024.
+Each monthly dataset is stored in Parquet format and accessed via Google Drive in a cloud-based Google Colab environment.
+We used Apache Spark (PySpark) for all data ingestion and transformation steps, given the dataset's scale of over 200 million records.
+
+\subsection{Data Processing}
+In PySpark, we cleaned the data by removing rows with null or invalid values (e.g., zero or negative durations and distances).
+We then created derived columns for \texttt{trip\_duration\_min}, \texttt{avg\_speed\_mph}, and categorical features like \texttt{pickup\_hour}, \texttt{pickup\_day}, and \texttt{trip\_provider}.
+Outliers were filtered using interquartile range (IQR) thresholds for trip duration, distance, and speed.
+These pre-processing steps reduced noise and ensured reliable downstream analysis.
+
+\subsection{Exploratory Analysis}
+Using Spark aggregations, we examined average trip speed and duration by hour of day, day of week, and provider.
+This analysis helped identify temporal congestion patterns and informed the design of the modeling phase.
+
+\subsection{Predictive Modeling}
+The cleaned dataset was exported to R for statistical modeling.
+A multiple linear regression model was used to predict trip duration based on features such as trip distance, pickup hour, and provider.
+A logistic regression model was also developed to classify trips as congested or not, using a threshold on duration (e.g., top 10\% longest trips).
+Model evaluation included interpretation of coefficients, R-squared for linear regression, and accuracy and confusion matrix metrics for logistic regression.
+
+\subsection{Method Justification and Tools}
+Given the volume and granularity of the dataset, traditional single-machine tools like pandas or base R would not scale.
+Apache Spark enables efficient, distributed computation across the full dataset.
+R is used for its strong support of regression modeling and statistical interpretation.
+Together, these tools enable scalable, interpretable, and reproducible analysis.
 

### 2025-04-20 paper.tex
chore: Draft paper.

@@
 Each monthly dataset is stored in Parquet format and accessed via Google Drive in a cloud-based Google Colab environment.
-We used Apache Spark (PySpark) for all data ingestion and transformation steps, given the dataset's scale of over 200 million records.
+We used Apache Spark (PySpark) for all data ingestion and transformation steps, given the dataset's scale of 239,470,448 records.
 
@@
 These pre-processing steps reduced noise and ensured reliable downstream analysis.
+Lastly, columns not needed for the analyses were removed from the dataset.
+The cleaned dataset contained nine columns compared to 24 in the original dataset.
 
@@
 
+\section{Results}
+\subsection{Descriptive Statistics and Temporal Trends}
+
+After pre-processing, the dataset contained approximately 204 million valid trip records.
+Average trip duration was 18.2 minutes with a standard deviation of 12.3 minutes, while average trip distance was 4.8 miles.
+Filtering removed extreme outliers, such as trips with durations over 120 minutes or speeds exceeding 80 mph.
+
+Temporal analysis showed that congestion peaks during weekday morning (7–9 AM) and evening (4–7 PM) hours, with significantly lower average speeds and longer trip durations.
+Weekends exhibited flatter congestion patterns with slightly shorter average trip times.
+Among providers, Lyft and Uber represented the majority of trips, and their congestion profiles closely mirrored citywide averages.
+
+\subsection{Regression Analysis}
+A multiple linear regression model was developed using trip distance, pickup hour, and provider as predictors of trip duration.
+The model achieved an R-squared value of 0.62, indicating a moderate fit.
+All predictors were statistically significant (p < 0.01), with trip distance contributing the most to variance explained.
+
+A logistic regression model classified trips as congested if their duration fell in the top 10% (greater than 37.5 minutes).
+The model achieved 86% accuracy with a precision of 0.79 and recall of 0.81.
+Pickup hour and provider were strong predictors, confirming temporal and provider-specific variation in congestion likelihood.
+
+These findings support the conclusion that ride-hailing congestion in NYC exhibits strong temporal regularities and that basic trip features can reliably forecast trip performance and congestion risk.
+
 \section{Conclusion}

### 2025-04-20 paper.tex
chore: Draft paper.

@@
 Lastly, columns not needed for the analyses were removed from the dataset.
-The cleaned dataset contained nine columns compared to 24 in the original dataset.
+The cleaned dataset contained five columns of the 24 in the original dataset (not including the aforementioned derived columns).
 
@@
 
-After pre-processing, the dataset contained approximately 204 million valid trip records.
+After pre-processing, the dataset contained 211,743,856 valid trip records.
 Average trip duration was 18.2 minutes with a standard deviation of 12.3 minutes, while average trip distance was 4.8 miles.
@@
 Weekends exhibited flatter congestion patterns with slightly shorter average trip times.
-Among providers, Lyft and Uber represented the majority of trips, and their congestion profiles closely mirrored citywide averages.
+Among providers, Lyft and Uber represented all trips in the dataset, and their congestion profiles closely mirrored citywide averages.
 
@@
 
-A logistic regression model classified trips as congested if their duration fell in the top 10% (greater than 37.5 minutes).
-The model achieved 86% accuracy with a precision of 0.79 and recall of 0.81.
+A logistic regression model classified trips as congested if their duration fell in the top 10\% (greater than 37.5 minutes).
+The model achieved 86\% accuracy with a precision of 0.79 and recall of 0.81.
 Pickup hour and provider were strong predictors, confirming temporal and provider-specific variation in congestion likelihood.

### 2025-04-21 paper.tex

@@
 \section{Abstract}
-This paper explores traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services such as Uber and Lyft.
-Using data from all 12 months of 2024, we investigate how trip durations and speeds vary by time of day, day of week, and provider.
-After extensive pre-processing in Apache Spark (PySpark), we identify temporal trends and build predictive models in R using linear and logistic regression.
+This paper explores traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services like Uber and Lyft.
+Using data from all 12 months of 2024, I investigate how trip durations and speeds vary by time of day, day of week, and provider.
+After extensive preprocessing in Apache Spark, I identify temporal trends and build predictive models in R using linear and logistic regression.
 The study demonstrates that the dataset size and analysis complexity necessitate big data tools for processing and modeling at scale.
@@
 Urban traffic congestion continues to be a pervasive problem affecting the efficiency of transportation systems, contributing to environmental concerns, and impacting commuter well-being.
-In densely populated cities like New York, where ride-hailing services such as Uber and Lyft operate at massive scale, understanding when and where congestion occurs is critical for optimizing urban mobility.
-This project investigates temporal patterns in congestion using High-Volume For-Hire Vehicle (FHV) trip data collected and published by the NYC Taxi and Limousine Commission (TLC).
+In densely populated cities like New York, where ride-hailing services such as Uber and Lyftoperate at massive scale, understanding when and where congestion occurs is critical for optimizing urban mobility.
+
+In this project, I investigate temporal patterns in congestion using High-Volume For-Hire Vehicle (FHV) trip data collected and published by the NYC Taxi and Limousine Commission (TLC).
 With over 200 million trip records from 2024 alone, the dataset provides a rich source of behavioral and operational insights.
 The central research question is: How do trip duration and speed, as proxies for congestion, vary across different times of day, days of the week, and ride-hailing providers?
-The objectives of the study are to (1) identify temporal congestion trends through descriptive analysis, and (2) develop regression models to predict trip duration or classify trips as congested based on their features.
+
+The objectives of this study are to (1) identify temporal congestion trends through descriptive analysis, and (2) develop regression models to predict trip duration or classify trips as congested based on their features.
 Given the volume and granularity of the dataset, traditional analysis tools fall short, necessitating the use of big data processing technologies like Apache Spark.
-This research aims to produce interpretable insights that can inform urban planning, traffic policy, and ride-hailing operations.
+My goal is to produce interpretable insights that can inform urban planning, traffic policy, and ride-hailing operations.
 
 \section{Background Review}
+Recent work by Riley et al.
+proposes an integrated optimization and machine learning framework to improve real-time ride-sharing dispatch using NYC trip data, reducing average passenger wait times by over 50\% \cite{riley2020dispatching}.
+Similarly, Ma et al.
+introduce a congestion-aware dispatch and charging scheduling system for electric ride-hailing fleets, achieving improved profitability and service efficiency in high-congestion scenarios \cite{ma2024congestionaware}.
+These studies highlight advanced, data-driven approaches to mitigating ride-hailing congestion and support the potential for predictive modeling grounded in large-scale NYC mobility data.
 
@@
 NYC’s Taxi and Limousine Commission has provided open access to large-scale trip data, which has enabled a range of analyses from temporal ridership patterns to spatial clustering of pick-up and drop-off zones \cite{nyctlc2023}.
+
 Several works have leveraged this dataset to evaluate ride-hailing’s impact on traffic flow and public transit usage.
-For example, a study by NYU's Tandon School of Engineering found that New York City’s congestion surcharge reduced the number of ride-hailing trips but did not significantly alleviate congestion \cite{nyu2023}.
+For example, a study by NYU Tandon School of Engineering found that New York City’s congestion surcharge reduced the number of ride-hailing trips but did not significantly alleviate congestion \cite{nyu2023}.
 Additionally, research from the CUNY Graduate Center examined the 2019 congestion surcharge’s effects on FHV usage in Manhattan, showing a decline in rides from charged zones \cite{cuny2023}.
+
 Other studies have investigated surge pricing’s influence on driver behavior \cite{surgepricing2022} and analyzed pay disparities across platforms using data from 2019 to 2022 \cite{ucla2023}.
 However, many of these analyses are limited in scope—often examining only a few months of data, focusing exclusively on yellow taxi or green cab datasets, or lacking predictive components.
+
 There remains a gap in literature concerning sustained, full-year temporal analysis of congestion based specifically on High-Volume FHV data.
 Furthermore, predictive modeling using such data is less common, particularly models focused on estimating congestion or trip delays.
-Our project addresses these shortcomings by utilizing a complete 12-month dataset and incorporating regression modeling.
-Additionally, our use of big data processing tools distinguishes this study from prior work that relies on single-machine or static data approaches.
+My project addresses these shortcomings by utilizing a complete 12-month dataset and incorporating regression modeling.
+Additionally, my use of big data processing tools distinguishes this study from prior work that relies on single-machine or static data approaches.
 
 \section{Method}
-This study follows a multi-phase approach consistent with the big data lifecycle, including data acquisition, pre-processing, analysis, and modeling.
+This study follows a multi-phase approach consistent with the big data lifecycle, including data acquisition, preprocessing, analysis, and modeling.
 
 \subsection{Data Acquisition and Setup}
-We obtained the NYC TLC High-Volume FHV trip record dataset for all twelve months of 2024.
+I obtained the NYC TLC High-Volume FHV trip record dataset for all twelve months of 2024.
 Each monthly dataset is stored in Parquet format and accessed via Google Drive in a cloud-based Google Colab environment.
-We used Apache Spark (PySpark) for all data ingestion and transformation steps, given the dataset's scale of 239,470,448 records.
+I used Apache Spark (PySpark) for all data ingestion and transformation steps, given the dataset's scale of over 200 million records.
 
 \subsection{Data Processing}
-In PySpark, we cleaned the data by removing rows with null or invalid values (e.g., zero or negative durations and distances).
-We then created derived columns for \texttt{trip\_duration\_min}, \texttt{avg\_speed\_mph}, and categorical features like \texttt{pickup\_hour}, \texttt{pickup\_day}, and \texttt{trip\_provider}.
+In PySpark, I cleaned the data by removing rows with null or invalid values (e.g., zero or negative durations and distances).
+I then created derived columns for trip_duration_min, avg_speed_mph, and categorical features like pickup_hour, pickup_day, and trip_provider.
 Outliers were filtered using interquartile range (IQR) thresholds for trip duration, distance, and speed.
-These pre-processing steps reduced noise and ensured reliable downstream analysis.
-Lastly, columns not needed for the analyses were removed from the dataset.
-The cleaned dataset contained five columns of the 24 in the original dataset (not including the aforementioned derived columns).
+These preprocessing steps reduced noise and ensured reliable downstream analysis.
+After preprocessing, the dataset contained approximately 204 million valid trip records.
+Average trip duration was 18.2 minutes with a standard deviation of 12.3 minutes, while average trip distance was 4.8 miles.
+Filtering removed extreme outliers, such as trips with durations over 120 minutes or speeds exceeding 80 mph.
 

### 2025-04-21 paper.tex
chore: Draft paper.

@@
-\documentclass{article}
+\documentclass[12pt,letterpaper]{article}
 \usepackage{graphicx} % Required for inserting images
@@
 \usepackage{xcolor}
-\usepackage{geometry}
 \usepackage{parskip}
@@
 \usepackage{url}
-
-\geometry{
- letterpaper,
- left=20mm,
- top=20mm,
- }
+\usepackage{setspace}
+\onehalfspacing
+\usepackage[margin=1in]{geometry}
 
@@
 
-\title{\vspace{-2cm}Temporal Congestion Analysis and Predictive Modeling of Ride-Hailing Trips in NYC Using High-Volume FHV Data}
+\title{{\vspace{-2cm}}Temporal Congestion Analysis and Predictive Modeling of Ride-Hailing Trips in NYC Using High-Volume FHV Data}
 \author{Farooq Mahmud}
@@
 \section{Abstract}
-This paper explores traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services like Uber and Lyft.
-Using data from all 12 months of 2024, I investigate how trip durations and speeds vary by time of day, day of week, and provider.
-After extensive preprocessing in Apache Spark, I identify temporal trends and build predictive models in R using linear and logistic regression.
-The study demonstrates that the dataset size and analysis complexity necessitate big data tools for processing and modeling at scale.
+This study outlines a potential project to explore traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services like Uber and Lyft.
+Using publicly available data from all twelve months of 2024, the proposed analysis would examine how trip durations and speeds vary by time of day, day of week, and provider.
+After preprocessing in Apache Spark (PySpark), I plan to identify temporal trends and build predictive models in R using linear and logistic regression.
+The project is designed to demonstrate how the size and complexity of the dataset require the use of big data tools for processing and modeling.
 
 \section{Introduction}
-Urban traffic congestion continues to be a pervasive problem affecting the efficiency of transportation systems, contributing to environmental concerns, and impacting commuter well-being.
-In densely populated cities like New York, where ride-hailing services such as Uber and Lyftoperate at massive scale, understanding when and where congestion occurs is critical for optimizing urban mobility.
+Urban traffic congestion continues to be a widespread problem that affects the efficiency of transportation systems, contributes to environmental concerns, and affects commuter well-being.
+In densely populated cities like New York, where ride-hailing services such as Uber and Lyft operate on a massive scale, understanding when and where congestion occurs is critical to optimizing urban mobility.
 
-In this project, I investigate temporal patterns in congestion using High-Volume For-Hire Vehicle (FHV) trip data collected and published by the NYC Taxi and Limousine Commission (TLC).
-With over 200 million trip records from 2024 alone, the dataset provides a rich source of behavioral and operational insights.
-The central research question is: How do trip duration and speed, as proxies for congestion, vary across different times of day, days of the week, and ride-hailing providers?
+This proposed project will investigate temporal patterns in congestion using High-Volume For-Hire Vehicle (FHV) trip data collected and published by the NYC Taxi and Limousine Commission (TLC).
+With more than 200 million travel records from 2024 alone, the dataset provides a rich source of behavioral and operational insights.
+The central research question is \textit{How do trip duration and speed, as proxies of congestion, vary between different times of the day, days of the week, and ride-hailing providers?}
 
-The objectives of this study are to (1) identify temporal congestion trends through descriptive analysis, and (2) develop regression models to predict trip duration or classify trips as congested based on their features.
-Given the volume and granularity of the dataset, traditional analysis tools fall short, necessitating the use of big data processing technologies like Apache Spark.
-My goal is to produce interpretable insights that can inform urban planning, traffic policy, and ride-hailing operations.
+The objectives of this project are to (1) identify temporal congestion trends through descriptive analysis and (2) develop regression models to predict the duration of the trip or classify the trips as congested based on their characteristics.
+Given the volume and granularity of the dataset, traditional analysis tools fall short, necessitating the use of big data processing technologies such as Apache Spark.
+The goal is to produce interpretable insights that can inform urban planning, traffic policy, and ride-hailing operations.
 
@@
 Recent work by Riley et al.
-proposes an integrated optimization and machine learning framework to improve real-time ride-sharing dispatch using NYC trip data, reducing average passenger wait times by over 50\% \cite{riley2020dispatching}.
+(2020) proposes an integrated optimization and machine learning framework to improve real-time ride sharing dispatch using NYC trip data, reducing average passenger wait times by more than 50\% \cite{riley2020dispatching}.
 Similarly, Ma et al.
 introduce a congestion-aware dispatch and charging scheduling system for electric ride-hailing fleets, achieving improved profitability and service efficiency in high-congestion scenarios \cite{ma2024congestionaware}.
-These studies highlight advanced, data-driven approaches to mitigating ride-hailing congestion and support the potential for predictive modeling grounded in large-scale NYC mobility data.
+These studies highlight advanced data-driven approaches to mitigating ride-hailing congestion and support the potential for predictive modeling grounded in large-scale NYC mobility data.
 
-A growing body of research has explored the use of ride-hailing data for understanding urban transportation dynamics.
+A growing body of research has explored the use of ride-hailing data to understand urban transportation dynamics.
 Previous studies have focused on topics such as demand prediction, ride-sharing optimization, pricing strategy, and transportation equity.
-NYC’s Taxi and Limousine Commission has provided open access to large-scale trip data, which has enabled a range of analyses from temporal ridership patterns to spatial clustering of pick-up and drop-off zones \cite{nyctlc2023}.
+The NYC Taxi and Limousine Commission has provided open access to large-scale trip data, which has allowed a variety of analyses, from temporal ridership patterns to spatial clustering of pick-up and drop-off zones \cite{nyctlc2023}.
 
-Several works have leveraged this dataset to evaluate ride-hailing’s impact on traffic flow and public transit usage.
-For example, a study by NYU Tandon School of Engineering found that New York City’s congestion surcharge reduced the number of ride-hailing trips but did not significantly alleviate congestion \cite{nyu2023}.
-Additionally, research from the CUNY Graduate Center examined the 2019 congestion surcharge’s effects on FHV usage in Manhattan, showing a decline in rides from charged zones \cite{cuny2023}.
+Several works have used this data set to assess the impact of ride-hailing on traffic flow and use of public transit.
+For example, a study by the NYU Tandon School of Engineering found that the New York City congestion surcharge reduced the number of ride-hailing trips, but did not significantly alleviate congestion \cite{nyu2023}.
+Furthermore, research from the CUNY Graduate Center examined the effects of the 2019 congestion surcharge on FHV usage in Manhattan, showing a decline in rides in charged zones \cite{cuny2023}.
 
-Other studies have investigated surge pricing’s influence on driver behavior \cite{surgepricing2022} and analyzed pay disparities across platforms using data from 2019 to 2022 \cite{ucla2023}.
-However, many of these analyses are limited in scope—often examining only a few months of data, focusing exclusively on yellow taxi or green cab datasets, or lacking predictive components.
+Other studies have investigated the influence of surge pricing on driver behavior \cite{surgepricing2022} and analyzed pay disparities between platforms using data from 2019 to 2022 \cite{ucla2023}.
+However, many of these analyses are limited in scope: often examining only a few months of data, focusing exclusively on yellow taxi or green cab datasets, or lacking predictive components.
 
-There remains a gap in literature concerning sustained, full-year temporal analysis of congestion based specifically on High-Volume FHV data.
+There remains a gap in the literature regarding sustained, full-year temporal analysis of congestion based specifically on High-Volume FHV data.
 Furthermore, predictive modeling using such data is less common, particularly models focused on estimating congestion or trip delays.
-My project addresses these shortcomings by utilizing a complete 12-month dataset and incorporating regression modeling.
-Additionally, my use of big data processing tools distinguishes this study from prior work that relies on single-machine or static data approaches.
+This project seeks to address these shortcomings using a complete 12 month dataset and incorporating regression modeling.
+Furthermore, the use of big data processing tools distinguishes this study from prior work that relies on single-machine or static data approaches.
 
-\section{Method}
-This study follows a multi-phase approach consistent with the big data lifecycle, including data acquisition, preprocessing, analysis, and modeling.
+\section{Proposed Method}
+This project will follow a multiphase approach consistent with the big data lifecycle, including data acquisition, preprocessing, analysis, and modeling.
 
 \subsection{Data Acquisition and Setup}
-I obtained the NYC TLC High-Volume FHV trip record dataset for all twelve months of 2024.
-Each monthly dataset is stored in Parquet format and accessed via Google Drive in a cloud-based Google Colab environment.
-I used Apache Spark (PySpark) for all data ingestion and transformation steps, given the dataset's scale of over 200 million records.
+I plan to obtain the NYC TLC High-Volume FHV trip record data set for all 12 months of 2024.
+Each monthly dataset is stored in Parquet format and can be accessed via Google Drive in a cloud-based Google Colab environment.
+I will use Apache Spark (PySpark) for all data ingestion and transformation steps, given the dataset's scale of over 200 million records.
 
 \subsection{Data Processing}
-In PySpark, I cleaned the data by removing rows with null or invalid values (e.g., zero or negative durations and distances).
-I then created derived columns for trip_duration_min, avg_speed_mph, and categorical features like pickup_hour, pickup_day, and trip_provider.
-Outliers were filtered using interquartile range (IQR) thresholds for trip duration, distance, and speed.
-These preprocessing steps reduced noise and ensured reliable downstream analysis.
-After preprocessing, the dataset contained approximately 204 million valid trip records.
-Average trip duration was 18.2 minutes with a standard deviation of 12.3 minutes, while average trip distance was 4.8 miles.
-Filtering removed extreme outliers, such as trips with durations over 120 minutes or speeds exceeding 80 mph.
+Data preprocessing plays a critical role in reducing both storage and compute requirements, resulting in not only time savings but also significant cost savings in a cloud-based environment.
+It also reduces the cognitive load on users by minimizing the number of columns they need to interpret and provides a standardized clean dataset that users can reliably use as the foundation for further analysis.
+By filtering out invalid records and outliers early in the pipeline, I can minimize the volume of data that needs to be processed in downstream stages, leading to more efficient use of Spark resources.
+
+In PySpark, the data will be cleaned by removing rows with null or invalid values (e.g., zero or negative durations and distances).
+I will create derived columns for \texttt{trip\_duration\_min}, \texttt{avg\_speed\_mph}, and categorical features like \texttt{pickup\_hour}, \texttt{pickup\_day}, and \texttt{trip\_provider}.
+Outliers will be filtered using interquartile range (IQR) thresholds for duration, distance, and speed of the trip.
+These preprocessing steps are designed to reduce noise and ensure reliable downstream analysis.
+
+\subsection{Operationalization and Scalability}
+To transition from a one-off analysis to a production-ready system, I propose extending the workflow to support automation and real-time data ingestion.
+Automated preprocessing of monthly datasets can be achieved by designing a parameterized PySpark pipeline and scheduling it with tools such as Apache Airflow \cite{airflow2024}.
+This setup would automatically clean and transform the dataset for each new month after publication, improving efficiency and standardization.
+
+In addition, the architecture could be extended to support real-time ingestion and streaming analytics.
+By integrating tools such as Apache Kafka with Spark Structured Streaming \cite{structuredstreaming2024}, the system could handle continuous data feeds from ride-hailing APIs or city sensors.
+This would enable use cases such as live congestion monitoring, real-time alerting, and dynamic traffic optimization.
+Operationalizing the pipeline in this way improves scalability and real-world applicability for ongoing urban analytics.
 
 \subsection{Exploratory Analysis}
-Using Spark aggregations, we examined average trip speed and duration by hour of day, day of week, and provider.
-This analysis helped identify temporal congestion patterns and informed the design of the modeling phase.
+Using Spark aggregations, I plan to examine the average speed and duration of the trip by hour of day, day of week, and provider.
+This analysis will help identify temporal congestion patterns and inform the design of the modeling phase.
 
 \subsection{Predictive Modeling}
-The cleaned dataset was exported to R for statistical modeling.
-A multiple linear regression model was used to predict trip duration based on features such as trip distance, pickup hour, and provider.
-A logistic regression model was also developed to classify trips as congested or not, using a threshold on duration (e.g., top 10\% longest trips).
-Model evaluation included interpretation of coefficients, R-squared for linear regression, and accuracy and confusion matrix metrics for logistic regression.
+The cleaned dataset can be imported into R for statistical modeling.
+I plan to build a multiple linear regression model to predict the duration of the trip based on characteristics such as the distance traveled, pickup hour, and the provider.
+In addition, I will develop a logistic regression model to classify trips as congested if their duration falls to the top 10\%.
+These models will be evaluated using metrics such as R-squared, accuracy, precision, and recall.
 
@@
 Given the volume and granularity of the dataset, traditional single-machine tools like pandas or base R would not scale.
-Apache Spark enables efficient, distributed computation across the full dataset.
-R is used for its strong support of regression modeling and statistical interpretation.
-Together, these tools enable scalable, interpretable, and reproducible analysis.
+Apache Spark will enable efficient, distributed computation across the full dataset.
+R is proposed for its strong support for regression modeling and statistical interpretation.
+Together, these tools provide a scalable, interpretable, and reproducible workflow.
 
-\section{Results}
-\subsection{Descriptive Statistics and Temporal Trends}
+\subsection{Security Management}
+Security is an essential consideration when designing a big data pipeline that processes public and potentially sensitive transportation data.
+Although the NYC TLC FHV dataset is publicly available and anonymized, future integration with private or real-time APIs may introduce data privacy and access control concerns.
+To ensure secure handling of data, role-based access control and data encryption should be enforced both in transit and at rest.
+When operationalizing the workflow in cloud environments, secure access credentials should be managed using secrets management tools such as AWS Secrets Manager \cite{awsSecretsManager2025}.
+In addition, audit logging and monitoring should be implemented to detect unauthorized access or anomalous usage patterns.
+These measures are necessary to ensure that the system adheres to data governance standards and maintains user trust.
 
-After pre-processing, the dataset contained 211,743,856 valid trip records.
-Average trip duration was 18.2 minutes with a standard deviation of 12.3 minutes, while average trip distance was 4.8 miles.
-Filtering removed extreme outliers, such as trips with durations over 120 minutes or speeds exceeding 80 mph.
-
-Temporal analysis showed that congestion peaks during weekday morning (7–9 AM) and evening (4–7 PM) hours, with significantly lower average speeds and longer trip durations.
-Weekends exhibited flatter congestion patterns with slightly shorter average trip times.
-Among providers, Lyft and Uber represented all trips in the dataset, and their congestion profiles closely mirrored citywide averages.
-
-\subsection{Regression Analysis}
-A multiple linear regression model was developed using trip distance, pickup hour, and provider as predictors of trip duration.
-The model achieved an R-squared value of 0.62, indicating a moderate fit.
-All predictors were statistically significant (p < 0.01), with trip distance contributing the most to variance explained.
-
-A logistic regression model classified trips as congested if their duration fell in the top 10\% (greater than 37.5 minutes).
-The model achieved 86\% accuracy with a precision of 0.79 and recall of 0.81.
-Pickup hour and provider were strong predictors, confirming temporal and provider-specific variation in congestion likelihood.
-
-These findings support the conclusion that ride-hailing congestion in NYC exhibits strong temporal regularities and that basic trip features can reliably forecast trip performance and congestion risk.
+\subsection{Storage Management}
+Managing storage efficiently is critical when working with high-volume data such as NYC's FHV trip records, which can span hundreds of millions of rows annually.
+Choosing the right file formats, such as Parquet, for compression and efficient querying helps reduce I/O overhead and long-term storage costs.
+Partitioning data by month or provider can further accelerate processing in distributed computing frameworks like Apache Spark.
+In a production setting, storage solutions, such as Amazon S3, offer scalable and cost-effective options with integrated lifecycle policies to automate data archival or deletion \cite{awsS3}.
 
 \section{Conclusion}
-This study highlights the feasibility and value of using big data tools to analyze ride-hailing trip data for congestion insights.
-It confirms the need for scalable data processing when working with multi-year FHV datasets and shows that simple predictive models can effectively estimate congestion levels using temporal features.
-Future work could include incorporating spatial features, extending modeling to multi-year data, or testing real-time streaming analysis.
+This proposed study is designed to demonstrate the value of using big data tools to analyze ride-hailing trip data to gain insight into congestion.
+It anticipates showing that simple predictive models can effectively estimate congestion levels using temporal features, while highlighting the importance of scalable infrastructure for large-scale mobility datasets.
+In future work, I may incorporate spatial features, extend modeling to multiyear data, or explore real-time streaming analysis.
 

### 2025-04-22 paper.tex
chore: Draft paper.

@@
 \section{Background Review}
-Recent work by Riley et al.
-(2020) proposes an integrated optimization and machine learning framework to improve real-time ride sharing dispatch using NYC trip data, reducing average passenger wait times by more than 50\% \cite{riley2020dispatching}.
-Similarly, Ma et al.
-introduce a congestion-aware dispatch and charging scheduling system for electric ride-hailing fleets, achieving improved profitability and service efficiency in high-congestion scenarios \cite{ma2024congestionaware}.
+Riley et al.
+propose a scalable real-time ride-sharing architecture (A-RTRS) that integrates optimization, machine learning, and predictive model control for dynamic dispatching and vehicle relocation on idle.
+Their system uses time-batched optimization, historical demand forecasting, and zone-level repositioning to reduce average passenger wait times by up to 55\%.
+This architecture demonstrates how predictive and adaptive mechanisms can operate in large-scale, high-velocity transportation environments.
+\cite{riley2020dispatching}.
+Ma et al.
+present a congestion-aware optimization framework for electric ride-hailing fleets that jointly coordinates vehicle dispatching and charging operations.
+Their system includes a day-ahead charging plan based on time-of-use energy pricing and a reactive model that adapts to vehicle availability and charger congestion in real time.
+Evaluated using simulated NYC taxi data, their congestion-aware approach outperformed benchmark policies in both service rate and profit, demonstrating the impact of integrated vehicle and charging scheduling on operational efficiency.
+\cite{ma2024congestionaware}.
 These studies highlight advanced data-driven approaches to mitigating ride-hailing congestion and support the potential for predictive modeling grounded in large-scale NYC mobility data.
@@
 Several works have used this data set to assess the impact of ride-hailing on traffic flow and use of public transit.
-For example, a study by the NYU Tandon School of Engineering found that the New York City congestion surcharge reduced the number of ride-hailing trips, but did not significantly alleviate congestion \cite{nyu2023}.
+A study by the NYU Tandon School of Engineering empirically evaluated the impact of NYC’s ride-hailing congestion surcharge policy using a difference-in-differences (DiD) approach.
+Although the policy led to a significant reduction in travel demand - 11\% overall, with Lyft experiencing the greatest drop - it did not result in a significant reduction in congestion.
+The study highlights that travel reductions were more pronounced for short-distance rides and in areas with alternative transportation options such as subways and bike sharing.
+It also found that the policy disproportionately affected lower income riders, raising concerns about equity and policy effectiveness\cite{nyu2023}.
+Although the Tandon study provides valuable insight into policy impacts using historical data, it lacks a supporting big data system architecture to enable real-time monitoring or operationalization.
+In contrast, the architecture proposed in this paper, which combines automated batch processing, real-time ingestion, and predictive modeling, could support dynamic congestion policy evaluation and feedback loops.
+Such a system could enable policy makers to monitor the impact of surcharges or routing incentives as they occur, enabling more responsive and equitable management of urban mobility.
 Furthermore, research from the CUNY Graduate Center examined the effects of the 2019 congestion surcharge on FHV usage in Manhattan, showing a decline in rides in charged zones \cite{cuny2023}.
+Although insightful, the CUNY study does not propose a scalable system for continuous or predictive policy evaluation.
+The big data architecture in this project would enable real-time pickup and collision monitoring, allowing faster feedback on congestion interventions and even potential automation of surcharge adjustments based on live trends.
+This infrastructure-oriented approach supports both exploratory and operational use cases that go beyond the economic evaluation described in the CUNY study.
 
@@
 It also reduces the cognitive load on users by minimizing the number of columns they need to interpret and provides a standardized clean dataset that users can reliably use as the foundation for further analysis.
-By filtering out invalid records and outliers early in the pipeline, I can minimize the volume of data that needs to be processed in downstream stages, leading to more efficient use of Spark resources.
+By filtering out invalid records and outliers early in the pipeline, I can minimize the volume of data that needs to be processed in downstream stages, leading to more efficient use of Spark and storage resources.
 
@@
 This setup would automatically clean and transform the dataset for each new month after publication, improving efficiency and standardization.
+Figure \ref{fig:big-data-architecture} outlines a potential production-level architecture to support batch ingestion, scalable processing, secure storage, and model analysis.
 
@@
 This would enable use cases such as live congestion monitoring, real-time alerting, and dynamic traffic optimization.
-Operationalizing the pipeline in this way improves scalability and real-world applicability for ongoing urban analytics.
+Operationalizing the pipeline in this way improves scalability and real-world applicability for continuous analytics.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{images/big-data-architecture.png}
+\caption{Production-level architecture}
+\label{fig:big-data-architecture}
+\end{figure}
 

### 2025-04-22 paper.tex
chore: Draft paper.

@@
 \subsection{Data Acquisition and Setup}
-I plan to obtain the NYC TLC High-Volume FHV trip record data set for all 12 months of 2024.
+I plan to obtain the NYC TLC High-Volume FHV trip record data set for all 12 months of 2024 \cite{dataset}.
 Each monthly dataset is stored in Parquet format and can be accessed via Google Drive in a cloud-based Google Colab environment.
@@
 This proposed study is designed to demonstrate the value of using big data tools to analyze ride-hailing trip data to gain insight into congestion.
-It anticipates showing that simple predictive models can effectively estimate congestion levels using temporal features, while highlighting the importance of scalable infrastructure for large-scale mobility datasets.
+It anticipates showing that simple predictive models can effectively estimate congestion levels using temporal features, while highlighting the importance of scalable infrastructure for large-scale datasets.
 In future work, I may incorporate spatial features, extend modeling to multiyear data, or explore real-time streaming analysis.

### 2025-04-22 paper.tex
chore: Draft paper.

@@
 
+\subsection{Limitations and Challenges}
+While this proposed methodology uses scalable technologies and sound modeling techniques, there are several limitations to consider.
+First, the dataset does not include contextual features, such as weather conditions, traffic incidents, or public transit disruptions, which can influence congestion and travel duration.
+Second, logistic regression assumes linearity between predictors and the log-odds of the outcome, which may oversimplify the relationship between variables such as time of day or provider and congestion.
+Third, the use of static, historical data limits the system's ability to capture sudden shifts in rider behavior, service availability, or city policy.
+Additionally, any real-time extension would depend on reliable APIs or streaming data access from service providers, which may not be publicly available.
+Finally, ensuring fairness and mitigating bias in modeling outcomes will be essential, especially when interpreting differences between providers or geographic regions.
+
 \section{Conclusion}

### 2025-04-22 paper.tex
chore: Draft paper.

@@
 \section{Abstract}
-This study outlines a potential project to explore traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from services like Uber and Lyft.
+This study outlines a potential project to explore traffic congestion patterns in New York City by analyzing High-Volume For-Hire Vehicle (FHV) trip records from ride-sharing services such as Uber and Lyft.
 Using publicly available data from all twelve months of 2024, the proposed analysis would examine how trip durations and speeds vary by time of day, day of week, and provider.
